---
title: "Oct 1 Activities: M&Ms; Survey; Sampling"
date: 9-30-2025
---
Ryan Bowie

All three class activities are on this .rmd notebook

Activity #1: Activity: M&Ms - Central Limit Theorem

Working in groups, each group will get a handful of candy from the bag, and then count the number of red and black candies. Log those in this Google Sheet 

https://docs.google.com/spreadsheets/d/1KOeoVTq2W7L2T9sWjwh5U_2f59GsTeAylVGD53rNWww/edit?usp=sharing

Rainbows

Take a look at the values that we all get - what do you notice about the patterns? This is an exercise about the central limit theorem. 

The central limit theorem states that the sampling distribution of the mean of any independent, random variable will be normal or nearly normal, if the sample size is large enough.

Then discuss and try to answer these questions:

--How close is your group's mean to the true proportion of 0.5?
--Why do different groups get different results?
--What happens to individual extreme results when we combine all groups?
--How does this relate to polls and surveys in real life?

Write your answer below. And then you may eat the M&Ms.

**A:**

Our mean is 86%, so we are somewhat far away from the true proportion of 0.5. Different groups get different results because they have different numbers of M&M's with different varieties. When we combine extreme results from all groups, our mean essentially lowers the overall mean. Most groups were closer to 90% than we were. In terms of polls, the same can be true when polling from a highly dense area vs a less populated area. 
------------------------------------------------------------------------------

# Activity #2:

Belonging Survey

Working in pairs, take a look back at the main findings from UMD's Belonging Survey:

https://president.umd.edu/internal/belonging


Specifically, focus on the "Sample Description" section and the Appendix. Discuss the differences and between the population and the sample, and what the document says about how the survey attempted to deal with them. Are there any decisions described in the document that raise questions in your minds? Detail the differences you see and your understanding of the methods they employed (it's ok if you don't understand all of them), and add your questions. You should discuss this as a pair but submit individually.

**A:**

Looking at the population vs samples it looks like men were undersampled and women were oversampled. One question that comes to mind is: What were the demographics of faculty responses and were they representative of the population? and Who was surveyed predominantly, students or faculty?

------------------------------------------------------------------------------

# TASK 3: SAMPLING

Based on the class sampling demo (available on Elms), sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

**A**
```{r}
library(tidyverse)

```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80

**A**
```{r}
set.seed(42)
age_ranges <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)

head(age_ranges)

```

### Task 3: Sample 100 rows of data from the dataset
**A**
```{r}
sample_100 <-age_ranges |> 
  sample_n(100)

```
### Lower Sample Size for Task 6
```{r}
sample_50 <- age_ranges |>
  sample_n(50)

```

### Task 4: Calculate the mean age and standard deviation of the sample
**A**
```{r}
sample_100 |> summarize(mean = mean(age), median = median(age), sd = sd(age))

```
### Mean and standard deviation for lower sample size
```{r}
sample_50 |> summarize(mean = mean(age), median = median(age), sd = sd(age))

```

### Task 5: Calculate the mean age and standard deviation of the population
**A**
```{r}
age_ranges |> summarize(mean = mean(age), median = median(age), sd = sd(age))

```


### Task 6: Compare the sample and population statistics
**A**

The mean is relatively similar between the mean of the sample and population, however the median is 50 in the population and 46.5 in the sample. The standard deviation is slightly higher in the sample than the population. Overall, the sample is representative of the population with no major outliers indicated. 


Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

**A**:

The means and standard deviations are similar with the sample size of 100, with a sample size of 50 the standard deviation is less than the population standard deviation and the median is slightly lower as well. The mean is still relatively consistent. Overall the results don't change much with a lower sample size.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
