---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries_2024 <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv") 
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

Most of the salaries are evenly distributed and gradually increase in a normal way. However, there are 11 employees with a salary of 0 and one employee (Michelle Barrans) that makes substallially more than the employee listed before her. These outliers could skew the data and affect the mean average of the salaries as a whole.

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries_2024 |>
summarize(mean_2024_salaries = mean(salary),
median_2024_salaries = median(salary)
)


```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries_2024 |>
summarize(mean_2024_salaries = mean(salary),
median_2024_salaries = median(salary),
min_2024_salaries = min(salary),
max_2024_salaries = max(salary)
)
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean of this data represents the average amount of money made per year by an employee on the White House staff in 2024, the median represents the person who is the middle earner out of everyone on the White House staff. The mininum tells me nothing, considering there is no way someone working as a White House staffer is working for free, the maximum tells me who the highest earner is. The highest earner and the 11 staffers with $0 salaries are the outliers.
### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

Based on the data, the story I would write would be about why the 11 White House staffers salaries are reported as $0. The next step I would take would be to reach out to each one of those departments where these staffers allegedly work and ask why their salaries aren't reported even though they are public officials and their salary information should be reported truthfully to the public.
