---
title: "HW7_MandMs"
date: 10/1/2025
---
Ryan Bowie

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)

```


### Task 2: Load an M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
candy <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```


### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10. Be sure to use set.seed(1234) before you start.
```{r}
set.seed(1234)

sample5 <- candy |>
  sample_n(5)

sample10 <- candy |>
  sample_n(10)

```


### Task 4: Calculate the mean, median and standard deviation of red M&Ms from each of the samples.
```{r}
sample5 |> summarize(mean = mean(red), median = median(red), sd = sd(red))

sample10 |> summarize(mean = mean(red), median = median(red), sd = sd(red))
```


### Task 5: Calculate the mean, median and standard deviation of red M&Ms from the population

```{r}
candy |> summarize(mean = mean(red), median = median(red), sd = sd(red))

```

### Task 6: Make three histograms: one each for sample5, sample10 and the original dataset. Include a headline. Use a consistent binwidth.

```{r}
sample5 |> 
  ggplot() + 
  geom_histogram(aes(x = red), binwidth = 2) +
labs(title = "Sample of 5 Rows Red M&M's ")

sample10 |> 
  ggplot() + 
  geom_histogram(aes(x = red), binwidth = 2) +
labs(title = "Sample of 10 Rows Red M&M's ")

candy |> 
  ggplot() + 
  geom_histogram(aes(x = red), binwidth = 2) +
labs(title = "Original Dataset Red M&M's ")

```


### Task 7: Answer questions


Compare the three sets of means, standard deviations and histograms. Which are more helpful to you in understanding the accuracy of the samples?

What does this exercise tell you about the relationship between sample size and estimate accuracy?

How does random sampling introduce variation into statistical estimates?

If you repeated this process, would the samples always be identical? Why or why not?


I feel like I did the histograms wrong, because they look pretty useless. The three sets of means is the most helpful in my opinion. All of the means are relatively similar to one another. The sample of 5 is 73.8, the sample of 10 is 74.7 and the population is 72.98. This suggests to me that all of the bags of M&Ms would have roughly a similar amount of red M&Ms only differing by one or two. If you sample more, you will have a more stable estimate that is closer to the population size. Random sampling can introduce variation, the number could be anything, but more samples will yield a better snapshot of what the true population is. No the samples wouldn't always be identical, they are random, they could be anything.


When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
