---
title: "Billboard Hot 100"
author: "Ryan Bowie"
date: "2025-11-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Opt+Cmd+I to create code chunk
```{r}
library(tidyverse)
```

```{r}
billboard <- read_csv("billboard.csv")
```
#STORY PITCH: I want to create a story about one-hit wonders and how they may or may not have diminished over the years.
```{r}
#Trying to make this more simplistic by making a table with just the songs, artist, date and weeks at number one

billboard_simple <- billboard |>
  select(Song, Artist, Date, `Weeks at Number One`) |>
  arrange(Date)
    

```
#The goal here is to summarize the 'Weeks at Number One' in order to figure out the amount of time that all of these songs have spent in the No.1 spot
```{r}

stats_weeks <- songs_year |>
  summarise(
    mean_weeks   = mean(`Weeks at Number One`, na.rm = TRUE),
    median_weeks = median(`Weeks at Number One`, na.rm = TRUE),
    min_weeks    = min(`Weeks at Number One`, na.rm = TRUE),
    max_weeks    = max(`Weeks at Number One`, na.rm = TRUE)
  )

#Through this, I figured out that on average, a song spends about 2.94 weeks in the No.1 spot with the most being at 19 and the least being at 1. 

```
#Now I will compile all songs with less than 3 weeks in the No.1 spot.

```{r}

songs_less_than_3_weeks <- billboard |>
  select(Song, Artist, Date, `Weeks at Number One`) |>
  filter(`Weeks at Number One` < 3)

```
#trying to separate the data into specific decades, no luck
```{r}
billboard <- billboard |>
  mutate(
    Year = as.integer(format(Date, '4-Aug-58':'11-Jan-25')),
    Decade = Year - (Year)
  ) # not working
```
```{r}
# working!!!!
library(dplyr)

billboard <- billboard %>%
  mutate(
    # grab the last two characters of the Date string (e.g., "58", "60")
    year_2 = as.integer(substr(Date, nchar(Date) - 1, nchar(Date))),
    # turn 58 -> 50, 60 -> 60, then make "50s", "60s"
    Decade = paste0(floor(year_2 / 10) * 10, "s")
  )
head(billboard)

billboard_simple <- billboard |>
  select(Song, Artist, Date, `Weeks at Number One`, Decade) |>
  arrange(Date)
   
```

#I hit a massive wall at this point and can't figure out where to go from here

#Analysis

This data contains all Billboard Hot 100 No.1 hits since 1958. 
The columns that I specifically want for this project is Song, Artist, Date and Weeks at Number One. I would like to make another column as well denoting what decade the songs are from to consolidate the data better.

Hypotheses:

Null hypothesis: One-hit wonder songs and multi-hit songs have the same average weeks at No.1

Alternative Hypothesis: One-hit wonder songs spend a different average number of weeks at No.1

#Can't go much further than this since I can't figure out the code above at this point. 


